---
title: "From default to publication-ready ggplot2"
subtitle: "A hands-on workshop"
author: "Selina Baldauf"
format:
  revealjs:
    slide-number: true
    code-line-numbers: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(gapminder)
library(scales)
library(patchwork)

# "Before" plots should show the ggplot2 default look
theme_set(theme_grey())

gap_2007 <- gapminder |> filter(year == 2007)

# The theme we build together in Module 1 (lives in solutions/theme.R)
theme_workshop <- function(base_size = 16, ink = "grey20", paper = "white") {
  theme_light(base_size = base_size, ink = ink, paper = paper) %+replace%
    theme(
      legend.text = element_text(size = rel(0.85)),
      panel.grid.minor = element_blank()
    )
}

# The recurring bubble chart (default colors, no theme layer)
p_bubble <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)
) +
  geom_point() +
  scale_x_log10(labels = label_dollar(accuracy = 1)) +
  scale_size(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    x = "GDP per capita (USD, log scale)",
    y = "Life expectancy (years)",
    color = "Continent",
    size = "Population"
  )
```

# Intro

## Default → publication-ready

A ggplot is easy to make. A *publication-ready* ggplot takes a few more deliberate choices.

So many things you can tune:

- The plot itself — which geoms to pick
- Themes, colours, fonts
- How to highlight your results
- Combining plots into panels
- Exporting at a readable size for the target medium

. . .

Different outputs want different tweaks — a **poster** or **talk** is not a **journal figure**.

## It's a rabbit hole

- ggplot2 has a whole ecosystem of extension packages for all of this
- You can spend endless time here

. . .

Today I'll show you a handful of **techniques and packages I find genuinely useful** — not an exhaustive tour, there's much more.

## Today

We'll focus on:

1. **Custom themes** — a consistent look, defined once
2. **Colour** — with intent, and colourblind-safe
3. **Directing attention** — guide the reader to the story
4. **Multi-panel layouts** — combine plots with `patchwork`
5. **Exporting** — at the right size and resolution

…plus a few other handy extension packages along the way.

## How it works

- There's a **GitHub repo** you download and open in your IDE (I'll use Positron)
- I demonstrate each topic — **follow along or just watch**, both are fine
- Each module has a short **exercise** to play with the concept
- **Questions** anytime: in the chat or just unmute

## Get set up

1. Click the green **Code** button → **Download ZIP**
2. Unzip it somewhere you can find it
3. Double-click the `.Rproj` file → opens in RStudio
   - *(Positron users: open the folder)*
4. Run `install_packages.R` to install today's packages

. . .

Repo: **[link in the chat]**

# Module 1: Custom themes

## The default isn't publication-ready

```{r}
#| echo: false
p_bubble
```

## Built-in themes: a quick win

Add a `theme_*()` layer and the whole look changes.

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5.5
mini <- function(th, name) {
  p_bubble +
    th +
    labs(title = name, x = NULL, y = NULL) +
    theme(legend.position = "none", plot.title = element_text(size = 13))
}
wrap_plots(
  list(
    mini(theme_grey(), "theme_grey()"),
    mini(theme_bw(), "theme_bw()"),
    mini(theme_light(), "theme_light()"),
    mini(theme_minimal(), "theme_minimal()"),
    mini(theme_classic(), "theme_classic()")
  ),
  nrow = 2
)
```

## Further customization with `theme()`

A theme controls everything that isn't data: Fonts, sizes, colours, grid lines, spacing, legends, ...

<iframe src="ggplot_theme_system_cheatsheet.pdf" width="100%" height="460" style="border:none;"></iframe>

::: {.aside}
ggplot2 theme-system cheatsheet (Henry Wang). Open the PDF in `slides/` for the full reference.
:::

## Inheritance of `theme` elements

```
text  ──┬── axis.text  ──┬── axis.text.x
        │                └── axis.text.y
        ├── plot.title
        ├── plot.subtitle
        └── legend.text

line  ──┬── axis.line   ──┬── axis.line.x
        │                 └── axis.line.y
        ├── axis.ticks
        └── panel.grid   ─── panel.grid.major / minor

rect  ──┬── plot.background
        └── panel.background
```

- Set a **parent** and it affects all its **children**
- E.g. Change `text` once and every piece of text updates

## How to change a theme

```{r}
p_bubble +
  theme_light(base_size = 18) +
  theme(
    legend.text = element_text(size = rel(0.85)),
    panel.grid.minor = element_blank()
  )
```

## Live demo

Open `R/01_themes.R` to follow along or just watch and wait for the exercise.

## Build a custom theme function

```{r}
#| eval: false
# Write a custom theme function using theme light as a basis
theme_workshop <- function(base_size = 16) {
  theme_light(base_size = base_size) %+replace%
    theme(
      legend.text = element_text(size = rel(0.85)),
      panel.grid.minor = element_blank()
    )
}

# Apply the custom theme
p_bubble + theme_workshop(base_size = 18)
```

## Put the theme in its own file

- Good practice: Save the theme function in its own file (e.g. `R/theme.R`) and `source()` it across scripts:

```r
source(here::here("R", "theme.R"))
p_bubble + theme_workshop()
```

## Set a global theme with `theme_set()`

Set the theme once at the top of your script (or in `theme.R`) and every plot in the session picks it up.

```r
# This theme will be used for every plot in the session
theme_set(theme_workshop())


# each plot uses theme_workshop() automatically
# no need to specify the theme again and again
p_bubble            
```

## Your turn

### Task (~5 min)

> Open `exercises/01_themes_exercise.R`

- Create a theme file `exercises/theme_custom.R`
- In it, write `theme_custom()` based on a built-in theme of your choice
- Source it in the exercise script and apply `theme_custom()` to `p_lines`

**Finished early?** Try different `base_size` values, or add more customizations.

Stuck? See `solutions/01_themes_final.R`.

# Module 2: Colour with intent

```{r}
#| label: m2-setup
#| include: false
library(scico)

okabe <- palette.colors(palette = "Okabe-Ito")
continent_colors <- c(
  Africa = okabe[[2]], Americas = okabe[[3]], Asia = okabe[[4]],
  Europe = okabe[[6]], Oceania = okabe[[1]]
)

# strip of discrete colours
swatch_discrete <- function(cols) {
  df <- data.frame(x = seq_along(cols))
  ggplot(df, aes(x, 1, fill = factor(x))) +
    geom_tile(width = 0.95) +
    scale_fill_manual(values = unname(cols), guide = "none") +
    theme_void()
}

# smooth strip from many colours
swatch_ramp <- function(cols) {
  ggplot(data.frame(x = seq_along(cols)), aes(x, 1, fill = x)) +
    geom_raster() +
    scale_fill_gradientn(colours = cols, guide = "none") +
    coord_cartesian(expand = FALSE) +
    theme_void() +
    theme(plot.margin = margin(2, 4, 2, 4))
}

# titled ramps in a grid (ncol controls layout)
ramp_panel <- function(named_cols, ncol = 1) {
  strips <- lapply(names(named_cols), function(nm) {
    swatch_ramp(named_cols[[nm]]) +
      labs(title = nm) +
      theme(plot.title = element_text(size = 12, hjust = 0))
  })
  patchwork::wrap_plots(strips, ncol = ncol)
}
```

## Pick intuitive colours

Use colours your reader already associates with the thing.

![[Blogpost on colours by Lisa Charlotte Muth (Datawrapper)](https://www.datawrapper.de/blog/colors)](intuitive_colors.png)


## Two things that matter

For scientific figures:

1. **Intuitive** colours that match what your reader expects
2. **Colourblind-safe** colours (about 1 in 12 men has a colour vision deficiency)

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 2.4
#| fig-align: center
set.seed(42)
cb_demo <- data.frame(
  x = runif(20, 0, 10),
  y = runif(20, 0, 5),
  group = rep(c("A", "B"), each = 10)
)
rg_cols <- c(A = "red", B = "forestgreen")
rg_deutan <- colorspace::deutan(rg_cols)
names(rg_deutan) <- names(rg_cols)

p_normal <- ggplot(cb_demo, aes(x, y, color = group)) +
  geom_point(size = 7, alpha = 0.9) +
  scale_color_manual(values = rg_cols, guide = "none") +
  labs(title = "Normal vision") +
  theme_void() +
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p_deutan <- p_normal +
  scale_color_manual(values = rg_deutan, guide = "none") +
  labs(title = "Deuteranopia")

p_normal + p_deutan
```

## Three kinds of palettes

:::: {.columns}

::: {.column width="33%"}
**Qualitative**

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_discrete(continent_colors)
```

Distinct categories
:::

::: {.column width="33%"}
**Sequential**

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_ramp(viridisLite::viridis(256))
```

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_discrete(viridisLite::viridis(7))
```

Ordered, low to high
:::

::: {.column width="33%"}
**Diverging**

```{r}
#| echo: false
#| fig-width: 3.2
#| fig-height: 0.7
swatch_ramp(scico::scico(256, palette = "vik"))
```

Above / below a midpoint
:::

::::

## Okabe-Ito (base R)

A colourblind-safe qualitative palette, built into base R.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 1
swatch_discrete(okabe)
```

- 8 distinct hues plus black and grey, separable under all common CVD types
- access with `palette.colors(palette = "Okabe-Ito")`


[See here for details](https://clauswilke.com/dataviz/color-pitfalls.html#fig:palette-Okabe-Ito)

## viridis (built into ggplot2)

Perceptually uniform, colourblind- and greyscale-safe. Best for ordered or continuous data.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.4
ramp_panel(list(
  magma   = viridisLite::viridis(256, option = "A"),
  inferno = viridisLite::viridis(256, option = "B"),
  plasma  = viridisLite::viridis(256, option = "C"),
  viridis = viridisLite::viridis(256, option = "D"),
  cividis = viridisLite::viridis(256, option = "E"),
  rocket  = viridisLite::viridis(256, option = "F"),
  mako    = viridisLite::viridis(256, option = "G"),
  turbo   = viridisLite::viridis(256, option = "H")
), ncol = 2)
```

- access with `scale_colour_viridis_d()` / `_c()` 

[See here for details](https://ggplot2.tidyverse.org/reference/scale_viridis.html)

## scico R package

Perceptually uniform and colourblind-safe. Made for **scientific data** and designed to be fair, readable, and citable


```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.4
ramp_panel(list(
  batlow  = scico::scico(256, palette = "batlow"),
  lajolla = scico::scico(256, palette = "lajolla"),
  bilbao  = scico::scico(256, palette = "bilbao"),
  oslo    = scico::scico(256, palette = "oslo"),
  vik     = scico::scico(256, palette = "vik"),
  roma    = scico::scico(256, palette = "roma"),
  berlin  = scico::scico(256, palette = "berlin"),
  broc    = scico::scico(256, palette = "broc")
), ncol = 2)
```

- access with `scale_colour_scico_d()` / `_c()`

[scico R package](https://github.com/thomasp85/scico) · [colour maps by Fabio Crameri](https://www.fabiocrameri.ch/colourmaps/)

## Check it: `cvdPlot()`

Simulates how your figure looks under common types of colour vision deficiency.

```{r}
#| output-location: slide
library(colorBlindness)
# Shrink text so the 4-panel cvdPlot output stays readable.
cvdPlot(p_bubble + theme(text = element_text(size = 8)))
```

see next slide for the result


## Live demo

Open `R/02_color.R`.

1. `cvdPlot()` the default, who can't read it?
2. Apply Okabe-Ito colours manually
3. A **named** colour vector for consistent continents
4. viridis and scico scale functions


## Your turn

### Task (~5 min)

> Open `exercises/02_color_exercise.R`

- Apply a colourblind-safe palette (viridis or scico) to the line chart
- Check it with `cvdPlot()`
- **If time:** build a named vector so each continent keeps its colour

Stuck? See `solutions/02_color_final.R`.

# Module 3: Directing attention

## Coming up

*(Slides to be built, see `workshop_notes.md`.)*

- From spaghetti to a single story
- `gghighlight`, `ggrepel`, `ggtext`, `annotate()`

# Module 4: Multi-panel layouts

```{r}
#| label: m4-setup
#| include: false
gap_continent <- gapminder |>
  group_by(year, continent) |>
  summarise(
    mean_lifeExp = mean(lifeExp),
    mean_gdpPercap = mean(gdpPercap),
    .groups = "drop"
  )

p_bubble_m4 <- ggplot(
  gap_2007,
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)
) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = label_dollar(accuracy = 1)) +
  scale_size(labels = label_number(scale_cut = cut_short_scale())) +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop(base_size = 16) +
  labs(
    x = "GDP per capita (USD, log scale)",
    y = "Life expectancy",
    color = "Continent",
    size = "Population"
  )

p_life <- ggplot(gap_continent, aes(year, mean_lifeExp, color = continent)) +
  geom_line(linewidth = 1) +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop(base_size = 16) +
  theme(legend.position = "none") +
  labs(
    y = "Life expectancy",
    color = "Continent",
    x = "Year"
  )

p_gdp <- ggplot(gap_continent, aes(year, mean_gdpPercap, color = continent)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = label_dollar(accuracy = 1)) +
  scale_color_scico_d(palette = "batlow") +
  theme_workshop(base_size = 16) +
  theme(legend.position = "none")+
  labs(
    y = "GDP per capita",
    color = "Continent",
    x = "Year"
  )
```

## Common multi-panel layouts

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6.5
p_bubble_m4 / (p_gdp + p_life) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A", tag_suffix = ")")
```

## The patchwork package

- Compose plots
- Collect shared legends and axes
- Tag panels
- Apply additional plot layers to every panel
- Place plots inside plots as insets

The [patchwork documentation](https://patchwork.data-imaginist.com/) is excellent

## Live demo

Open `R/04_patchwork.R`.

## Your turn

### Task (~5 min)

> Open `exercises/04_patchwork_exercise.R`


# Module 5: Exporting figures

## Coming up

*(Slides to be built, see `workshop_notes.md`.)*

- `ggsave()` dimensions and dpi
- `ggview()` preview, `base_size` for print, `ragg` / `cairo_pdf`
